---
title: "Data wrangling & visualization (Sierra amphibians)"
author: "Jaxon Stuhr"
date: "1/18/2022"
output: html_document
---

This code wrangles and visualizes data collected by the Sierra Lakes Inventory Project (SLIP). Frogs of a variety of species were observed in southern Sierra Nevada water bodies between 1995 and 2002. The following visualizations focus on mountain yellow-legged frogs (*Rana muscosa*) and sort by year, life-stage, and lake in which they were observed. 

```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(readxl)
library(tidyverse)
library(cowplot)
library(here)
library(lubridate)
```

```{r}
# read in data 
amphs_raw = read_excel(here("data", "sierra_amphibians.xlsx"))
# clean up dates and add "year" column
amphs = amphs_raw %>% 
  mutate(survey_date = ymd(survey_date)) %>% 
  mutate(year = year(survey_date))
```

```{r}
# create subset with only Adult, SubAdult, and Tadpole mountain yellow-legged frogs (RAMU)
amphs_sub_1 = amphs %>% 
  filter(
    amphibian_life_stage %in% c("Adult", "SubAdult", "Tadpole"),
    amphibian_species == "RAMU"
    )
# groupby year and life stage and summarize counts
amphs_summary_1 = amphs_sub_1 %>% 
  group_by(amphibian_life_stage, year) %>% 
  summarize(count = n())
```

```{r}
# plot count by year colored by life-cycle stage
ggplot(amphs_summary_1, aes(x = year, y = count, fill = amphibian_life_stage)) + 
  geom_col() +
  guides(fill=guide_legend(title="Life Stage")) + 
  labs(x = "Year", y = "Number Observed Frogs") + 
  theme_minimal()
```

Figure 1: Observed mountain yellow-legged frogs (*Rana muscosa*) by year and life stage in southern Sierra Nevada water bodies

```{r}
# create subset with only Adult, SubAdult, and Tadpole mountain yellow-legged frogs (RAMU)
amphs_sub_2 = amphs %>% 
  filter(
    amphibian_life_stage %in% c("Adult", "SubAdult"),
    amphibian_species == "RAMU"
    )
# group_by lake_id, slice top 5 counts, and rename lakes
amphs_summary_2 = amphs_sub_2 %>% 
  group_by(lake_id) %>% 
  summarize(count = n()) %>% 
  slice_max(order_by = count, n = 5) %>% 
  mutate(lake_id = case_when(
    lake_id == 50183 ~ "Lake 50183",
    lake_id == 70583 ~ "Lake 70583",
    lake_id == 10226 ~ "Lake 10226",
    lake_id == 41322 ~ "Lake 41322",
    lake_id == 50219 ~ "Lake 50219"
  )) %>% 
  # order by count
  mutate(lake_id = fct_reorder(lake_id, count))
```

```{r}
# plot frog counts by lake_id
ggplot(amphs_summary_2, aes(x = lake_id, y = count)) + 
  geom_col(fill = 'darkgreen') + 
  theme_minimal() + 
  labs(x = "Number Observed Frogs", y = "Number Observed Frogs")
```

Fig 2: Observed mountain yellow-legged frogs (*Rana muscosa*) for five southern Sierra lakes with most observations from 1995 - 2002

Data citation:

Knapp, R.A., C. Pavelka, E.E. Hegeman, and T.C. Smith. 2020. The Sierra Lakes Inventory Project: Non-Native fish and community composition of lakes and ponds in the Sierra Nevada, California ver 2. Environmental Data Initiative. https://doi.org/10.6073/pasta/d835832d7fd00d9e4466e44eea87fab3
